Supporting Information

### 2021-09-01 vertical-thermal-review.Rmd
properly cite all SI materials
@NidhiVinod , this is taken care of.

#SI table numbers
NEON_sites_legend = "S1"
Tealeaves_parameters = "S2"

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF). Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig. S1. Further site information is given in Supporting Information Table `r NEON_sites_legend`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`a-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types [Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020]. Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`. [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/fig3_leafenergy.png)

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).

>>>>

#SI table numbers
NEON_sites = "S1"
Tealeaves_parameters = "S2"

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S. National Ecological Observatory Network [NEON; Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF). Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig. S1. Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types [Fig. `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020]. Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).**
The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Fig. `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`. Model parameters for biophysical conditions are given in Supplementary Information Table `r Tealeaves_parameters`.](leaf_energy_balance/fig3_leafenergy.png)

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
